Quantum H theorem, variational entropy, and relaxation processes

### 2021-01-09 tex/quantum.tex
Rewording: up to sec. 3.2.1

In equilibrium (at $t\to\infty$), $dn_{\kappa}/dt=0$, hence from Eq.~(\ref{eq:changen})
%
\begin{equation}\label{eq:boseiferdirdistfnctsrc}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}}.
\end{equation}
%
Considering that the energy is conserved during the collsion, the Bose-Einstein or Fermi-Dirac distribution
functions can be recovered from Eq.~(\ref{eq:boseiferdirdistfnctsrc}):
%
\begin{equation}
   n_{\kappa}=\frac{g_{\kappa}}{\exp(\alpha+\beta\epsilon_{\kappa}\mp1)}.
\end{equation}
%

@@

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M\label{entropy}.
\end{eqnarray}
%
Here, the upper and lower sign refer to bosons and fermions respectively.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and its local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M
\end{equation}
\end{subequations}
%
which are, in addtion, constrained by the global microcanonical restrictions
%
\begin{subequations}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

@@

%============================================================
\subsubsection{Properties of $\mathcal{H}$ in equilibrium}
%============================================================
If the system is in equilibrium (at $t\to\infty$), the local number of particles and the local
energy do not depend on the cell number and they should be homogeneous, this is \commr{[This needs a rationale\dots]}
%
\begin{subequations}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K
\end{equation}
%
and
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K.
\end{equation}
%
\end{subequations}
%

@@

%
\commr{We can obtain
from the variational procedure that
%
\begin{equation}
    \alpha_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{N}_M}, \ \ \ \beta_M\propto \frac{\partial \ln \Omega_M}{\partial \mathcal{E}_M},\label{multipliers}
\end{equation}
%
[What is $\Omega_M$?]
}

where we used that the $H$-functional in equilibrium is proportional to the
entropy of the system.
In other hand, in the expression (\ref{multipliers}), we expect that the
variation of the logarithm with the number of particles by cells is indifferent
from a particular cell. With this, we conclude 
%
\begin{equation}
    \alpha_M=\alpha; \ \ \ \ \beta_M=\beta.
\end{equation}
%
In equilibrium, the energy and the Lagrange multiplier do not depend on the
cell number and time, therefore
%
\begin{equation}
    \bar f_{Mn}(\epsilon_{n},t)=\bar f_n(\epsilon_{n}) =\frac{1}{e^{-\alpha-\beta \epsilon_n}\pm 1}.
\end{equation}
%
If one compares those multipliers with the mean energy and the mean particle
number from the statistical mechanic's results, we obtain
%
\begin{equation}
    \alpha=\frac{\mu}{kT}\equiv \bar{\alpha}; \ \ \ \ \beta=-\frac{1}{kT}\equiv \bar{\beta},
\end{equation}
%
and finally
%
\begin{eqnarray}
    \bar{f}_{n}(\epsilon_{n})&=&\frac{1}{e^{(\frac{{\epsilon_n}-\bar{\mu}}{kT})}\pm 1}\equiv \bar{f}_{n}.
\end{eqnarray}
%
We proved, when the system is in equilibrium, the Lagrange multipliers don't
depend on the cell number. In other words, Lagrange multipliers are homogeneous
in the system.

We can calculate the entropy from an ideal system, in other words, if one
substitute (\ref{distributionequilibrium}) to (\ref{entropy}) we obtain
%
\begin{eqnarray}
      \mathcal{H}(t)&=&\sum_{M=1}^{K} \sum_n  \left[\left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right)\ln \left(\frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}} \right) \right]\nonumber \\
      &&\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}\left(1 \mp \frac{1}{e^{-\bar{\alpha}-\bar{\beta}\epsilon_{n}}\pm 1} \right) \right] \de V_M \delta \epsilon_n\label{H-entropy}.
  \end{eqnarray}
%
This result allows us to calculate the entropy of a quantum ideal gases. This
entropy could be used as a thermodynamic variable to describe those systems.

%\subsection{The quantum analogous of the Boltzmann's $H$-theorem}
We rewrite the \textit{variational entropy} (\ref{entropy}) as
%
\begin{equation}
    \mathcal{H}(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.

>>>>

In equilibrium (at $t\to\infty$), $dn_{\kappa}/dt=0$, hence from Eq.~(\ref{eq:changen})
%
\begin{equation}\label{eq:boseiferdirdistfnctsrc}
    \ln \frac{n_{\kappa}}{g_{\kappa}\pm n_{\kappa}}+\ln \frac{n_{\lambda}}{g_{\lambda}\pm n_{\lambda}}=\ln \frac{n_{\mu}}{g_{\mu}\pm n_{\nu}}+\ln \frac{n_{\nu}}{g_{\nu}\pm n_{\nu}}.
\end{equation}
%
Considering that the energy is conserved during the collsion, the Bose-Einstein or Fermi-Dirac distribution
functions can be recovered from Eq.~(\ref{eq:boseiferdirdistfnctsrc}):
%
\begin{equation}
   n_{\kappa}=\frac{g_{\kappa}}{\exp(\alpha+\beta\epsilon_{\kappa})\mp1}.
\end{equation}
%

@@

We propose the following functional as an alternative $H$-functional
for quantum non-homogeneous diluted gases:
%
\begin{eqnarray}\label{eq:qHdef}
    \mathcal{H} (t)&=&\sum_{M=1}^{K} \sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)\bigg.\nonumber \\
    &&\bigg.\pm \Big(1 \mp f_{Mn}(t)) \ln (1 \mp f_{Mn}(t)\Big) \Big]   \delta V_M.
\end{eqnarray}
%
Here, the upper and lower sign refer to \commr{bosons and fermions respectively}.
In addition, when needed, each cell will have an associated local chemical potential,
$\alpha_M$, and its local $H$-functional, which is defined by:
%
\begin{equation}\label{eq:qHMdef}
   \mathcal{H}_M(t)=\sum_{n} \bigg[ f_{Mn}(t) \ln f_{Mn}(t)
   \pm \Big(1 \mp f_{Mn}(t)\Big) \ln \Big(1 \mp f_{Mn}(t)\Big) \bigg] \de V_M.
\end{equation}
%
Therefore, $\mathcal{N}_M$ and $\mathcal{E}_M$ as functions of time are given by:
%
\begin{subequations}
\begin{equation}
    {\mathcal{N}}_M(t) = \sum_{n}f_{Mn}(t) \delta V_M
\end{equation}
%
and
%
\begin{equation}
\mathcal{E}_M(t) = \sum_{n}f_{Mn}(t)\epsilon_{n} \de V_M
\end{equation}
\end{subequations}
%
which are, in addtion, constrained by the global microcanonical restrictions
%
\begin{subequations}
\begin{equation}\label{eq:qGlobRestN}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\right]\delta  V_M
    =\sum_{M=1}^{K}\mathcal{N}_{M}(t)\delta V_M=N
\end{equation}
%
and
%
\begin{equation}\label{eq:qGlobRestE}
    \sum_{M=1}^{K}\left[\sum_{n}f_{Mn}(t)\epsilon_{n}\right]\delta V_M
    =\sum_{M=1}^{K}\mathcal{E}_M(t)\delta V_M=E. 
\end{equation}
\end{subequations}
%

@@

%============================================================
\subsubsection{Properties of $\mathcal{H}$ in equilibrium}
%============================================================
If the system is in equilibrium (at $t\to\infty$), the temperature becomes homogeneous throughout
the complete system. Also, the local number of particles, the local
energy, and the Lagrange multipliers do not depend on the cell number and they should be homogeneous,
this is \commr{[This needs a rationale\dots]}
%
\begin{subequations}\label{eq:qEqRestrictions}
\begin{equation}
   {\mathcal{N}}_M(t\to\infty)\equiv \bar{\mathcal{N}}=N/K,
\end{equation}
%
%
\begin{equation}
	{\mathcal{E}}_M(t\to\infty)\equiv \bar{\mathcal{E}}=E/K,
\end{equation}
\end{subequations}
%
%
\begin{subequations}\label{eq:qEqAlphaBeta}
\begin{equation}
	\alpha_M=\bar\alpha,\quad\forall M,
\end{equation}
%
and
\begin{equation}
	\beta_M=\bar\beta, \quad\forall M.
\end{equation}
\end{subequations}
%
Substituting Eqs.~(\ref{eq:qEqAlphaBeta}) into Eq.~(\ref{eq:qfMn}) yields the distribution
function of each cell in equilibrium to be
%
\begin{equation}\label{eq:qfneq}
    \bar f_{Mn}=\bar f_n =\frac{1}{\exp\big(-\bar\alpha-\bar\beta \epsilon_n\big)\pm 1},\quad\forall M.
\end{equation}
%
By means of the above equation, we can recover the distribution function and the entropy of a diluted quantum gas in equilibrium as follows. Setting $\bar\alpha=\mu/kT$ and $\bar\beta=-1/kT$,
and substituting them in Eq.~(\ref{eq:qfneq}) renders the Fermi-Dirac and Bose-Einstein
distribution functions:
%
\begin{equation}
    \bar{f}_{n}=\frac{1}{\exp\big(\frac{{\epsilon_n}-\mu}{kT}\big)\pm 1},
\end{equation}
%
and substituting Eq.~(\ref{eq:qfneq}) into the negative of Eq.(\ref{eq:qHdef}) provides the entropy:
%
\begin{eqnarray}\label{eq:varEntropyDef}
      S&=&\sum_{M=1}^{K}
        \sum_n  \left[
        		\left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
           \commr{\ln \left(\frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)}
         \right]\nonumber \\
      &&\quad\pm  \ln \left[\prod_{M=1}^{K} \prod_{n}
         \left(1 \mp \frac{1}{\exp\big(-\bar{\alpha}-\bar{\beta}\epsilon_{n}\big)\pm 1} \right)
         \right] \delta V_M.
\end{eqnarray}
%
For obtaining the last result, we have used the well-known fact that the $H$-functional
is the negative of the entropy. The property provided by Eq.(\ref{eq:varEntropyDef})
is precisely what we call the ``variational entropy'',
and this name reflects the fact that it was obtained \textit{via}
the variational method.

%=============================================
\subsubsection{Proof of the quantum $H$-theorem for non-homogeneous systems}
%=============================================

@@

We rewrite the \textit{variational entropy} (\ref{eq:qHdef}) as
%
\begin{equation}
    \mathcal{H}(t)=\sum_{M=1}^{K} \sum_{n} \left[ f_{Mn} \ln f_{Mn} \pm (1 \mp f_{Mn}) \ln (1 \mp f_{Mn}) \right] \de V_M \delta \epsilon_n \label{entropy2},
\end{equation}
%
where, for simplicity, we have defined $f(\epsilon_{n},t)\equiv f_{Mn}$.
